An open-source full 3D electromagnetic modeler for 1D VTI media in Python: empymod

### 2017-06-03 ManuscriptGeophysics2016.tex
Figure adjustements for final submission

% \usepackage{draftwatermark}
% \SetWatermarkLightness{.95}
% \SetWatermarkFontSize{2cm}
% % \SetWatermarkText{\shortstack[c]{Submitted to \texttt{Geophysics}\\23 Oct 2012}}
% \SetWatermarkText{\shortstack[c]{Re-submitted to \texttt{Geophysics}\\05 Feb 2013}}

%% ~ ADDITIONAL PACKAGES TO GEOPHYSICS.CLS
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\usepackage{color}
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage[USenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
% \usepackage[pdftex, draft]{hyperref}
\usepackage{xspace}                 % For spaces after \newcommand-strings
\usepackage{upquote}

% vvv STUFF NOT FOR SUBMISSION
\usepackage[pdftex, final]{hyperref}
\definecolor{myblue}{rgb}{0,0,.5}
\hypersetup{allcolors=myblue, allbordercolors={0 0 .5},
            pdfborderstyle={/S/U/W .5}, colorlinks=true,}
% ^^^ STUFF NOT FOR SUBMISSION

@@

%
% Figure and table widths
%   SEGTex               column width = 250 pt
%   Geophysics           column width = 240 pt (3.33 in)
%   Geophysics       1.3 column width = 312 pt ( 4.33 in)
% \plot[btp]{figure1}{width=240pt}{Diff porosities higher than roughly 40~\%.}
% \plot*[btp]{figure11}{width=\textwidth}{(a) i horizontal resistivities.}
%
\definecolor{MyGray}{gray}{0.85}
\newcolumntype{Y}{>{\columncolor{MyGray}\raggedleft\arraybackslash}r}
\newcolumntype{W}{>{\raggedleft\arraybackslash}r}
\newcolumntype{A}{>{\columncolor{MyGray}\raggedleft\arraybackslash}X}
\newcolumntype{B}{>{\raggedleft\arraybackslash}X}

@@

\ms{GEO-2016-0626.R2}

@@

% 1.4 1D codes and licenses
In terms of freely available and open-source code there are a few examples.
\cite{GEO.09.Key} published his forward modeling and inversion code Dipole1D,
written in Fortran. The dipole can be placed anywhere in the stack of layers
and can have arbitrary orientation and dip. Dipole1D computes the electric and
magnetic fields to an electric source, using the FHT method. \cite{GEO.12.Key}
published additional code with his introduction of the
qua\-dra\-ture-with-ex\-tra\-po\-la\-tion method (QWE) and its comparison to
the FHT method, for which he translated some of the Dipole1D-Fortran code to
Matlab. \cite{GEO.15.Hunziker} published their code EMmod (Fortran and C), in
which the source and receiver can also be placed anywhere in the stack of
layers. They use a 61\,pt Gauss-Kronrod quadrature for the Hankel transform.

@@

To my knowledge, empymod is the first code that is freely available which
calculates the full wavefield for a layered-earth model with vertical isotropy
and has different types of Hankel transforms (QWE, QUAD, FHT) and various
methods for the Fourier transform (FFTLog, sine/cosine-filters, QWE, FFT). An
interesting use-case for empymod is therefore comparison studies of the
different methods, up to building hybrid inversion schemes. The code is
published under the lax permissive \emph{Apache Version 2.0 license}, which
makes it available to everyone for free, even for commercial purposes. It might
hence prove to be valuable for students and professionals alike, and it is
valuable for the objectives of reproducible research. It is written in Python,
a modern, cross-platform, free and open-source programming language. With its
scientific libraries, mainly the numeric and scientific modules \texttt{NumPy}
and \texttt{SciPy}, it creates an extremely powerful numerical calculation
stack: Even though it is an interpreted language it can be very fast, as
\texttt{NumPy}/\texttt{SciPy} use under the hood routines in Fortran or in C
(using for instance the BLAS/Lapack libraries) for the linear algebra
computations, and Python is merely the glue.  The comparisons show that empymod
is as fast as the existing codes. The code is vectorized (avoiding loops) as
much as possible, which improves computation.  The most time-consuming
calculations have furthermore a flag to run parallelized.  However, on a larger
scale such as an inversion routine one would probably want to parallelize the
kernel calls instead of the kernel itself. The code is hosted on GitHub, which
makes it easy for anyone to improve and contribute to the code.  It comes with
an extensive testing suite which is run automatically on each commit, which
guarantees some robustness and helps to avoid regression bugs. Installation
guidelines, documentation, and examples of its usage can be found on its
website at \url{https://empymod.github.io}.  \cite{TLE.17.Werthmuller} gives a
tutorial-style introduction to CSEM modeling using empymod.

@@

The wavenumber-domain calculation in \emph{kernel} follows
\cite{GEO.15.Hunziker}, and calculates as such the complete wavefield for a
layered VTI model. The code makes no assumptions about the model: Source and
receiver can be placed anywhere in the model, including first and last layer,
and you have to define if the first layer is air or not. Bipoles can cross
layer boundaries. Depths, frequencies, and source-receiver configuration have
to be defined, and each layer is characterized with its horizontal resistivity
$\rho_\textrm{h}$, its electrical anisotropy $\lambda$, where $\lambda =
\sqrt{\rho_\textrm{v}/\rho_\textrm{h}}$, its horizontal and vertical magnetic
permeabilities $\mu_\mathrm{h}$ and $\mu_\mathrm{v}$, and the horizontal and
vertical electric permittivities $\epsilon_\mathrm{h}$ and
$\epsilon_\mathrm{v}$. The main difference between EMmod and empymod is the
Hankel transform, and a few additional fundamental differences: EMmod uses 2nd
order Bessel functions of the first kind $J_2$. Published FHT filters
generally provide coefficients for 0th and 1st order Bessel functions $J_0,
J_1$ only. Consequently, the recurrence relation
%
\begin{equation}
  J_2(kr) = \frac{2}{kr}J_1(kr) - J_0(kr)
  \label{eq:j2}
\end{equation}
%
is used, where $k$ and $r$ are the wavenumber and space-domain parameters,
respectively. Other differences are that empymod includes Fourier transforms,
hence time-domain calculation, and can model arbitrary rotated, finite bipoles.

@@

The most import part of \cite{GEO.12.Key} is the introduction of a new
quadrature algorithm to geophysics, named
qua\-dra\-ture-with-ex\-tra\-po\-la\-tion (QWE). QWE is a fast quadrature
method using the Shanks transformation \citep{JMP.55.Shanks} computed with
Wynn's epsilon algorithm \citep{MC.56.Wynn}. The advantage of quadrature over
filters is the ability to estimate the error. QWE continues until the absolute
error, estimated by the difference of subsequent iterations $n$, satisfies the
inequality
%
\begin{equation}
  |S^*_n-S^*_{n-1}| \le \varepsilon_r|S^*_n| + \varepsilon_a\ ,
  \label{eq:err}
\end{equation}
%
where $S^*$ is the extrapolated result, and $\varepsilon_r, \varepsilon_a$ are
the relative and absolute tolerance, respectively.

Whereas for the Hankel transform the three methods QWE, FHT, and QUAD, a
standard adaptive quadrature using QAGSE from the Fortran QUADPACK library, are
implemented, for the Fourier transform there are four methods implemented: QWE,
FHT with the sine and cosine filters, the standard FFT, and FFTLog
\citep{RAS.00.Hamilton}. The logarithmic fast Fourier transform FFTLog is
ideal for this operation, as generally a wide range of frequencies is required
to go from the frequency to the time domain. The FFTLog can yield faster
results than the QWE method and more precise than the FHT results, specifically
for land impulse responses at early times.

Calculation time can be reduced by using the splined version of the QWE and the
splined and lagged versions of the FHT. In addition, all time-consuming
calculations are set up to be able to run in parallel, with the help of the
Python-module \texttt{numexpr}. This can significantly speed up calculations if
you run big models with many layers and for many offsets and frequencies.
However, if you include empymod in an inversion scheme then it might be better
to parallelize the calls to empymod, instead of empymod itself.

@@

The run-time comparisons tests were run on a Lenovo ThinkCentre running Ubuntu
16.04 64-bit, with 8\,GB of memory and an Intel Core i7-4770 CPU @ 3.40GHz x 8.
Comparing run times is always a difficult task, and between different languages
even more, here between Python (empymod), Fortran (EMmod, Dipole1D), and Matlab
(scripts from \cite{GEO.12.Key}). However, they do serve as comparison. I
used the Matlab-\texttt{timing} and Python-\texttt{timeit} functions, which
behave very similar. It is probably expected today, but still worth mentioning,
that the calculation is carried out in double precision.

@@

\cite{PIER.10.Slob} published analytical frequency- and time-domain solutions
for the diffusive electric field in a VTI half-space, where source and receiver
can be placed anywhere in the half-space. As an example, I use the same model
as \cite{GEO.15.Hunziker} in their Figures~1 and~2: A half-space with
horizontal resistivity $\rho_\mathrm{h} = 1/3\,\Omega$\,m, anisotropy $\lambda
= \sqrt{10}$, and for frequency $f = 0.5\,$Hz. The source is located
horizontally at the origin ($x_\mathrm{s} = y_\mathrm{s} = 0\,$m) at a depth of
150\,m, and the depth of the receivers is 200\,m. The field is calculated on a
regular grid with a spacing of 10\,m. Figure~\ref{fig:analytical} shows the
analytical amplitude and phase results for the first quadrant (the other
quadrants are simply symmetric copies of it).
%
\plot*[btp]{analytical}{width=\textwidth}{Analytical solution for a half-space
  with $\rho_\mathrm{h} = 1/3\,\Omega$\,m, $\lambda = \sqrt{10}$, $f =
  0.5\,$Hz, $z_\mathrm{s} = 150\,$m, and $z_\mathrm{r} = 200\,$m, calculated on
  a regular grid with a spacing of 10 meters; (a) amplitude and (b) phase.}
%
The figure shows the result for x-directed electric source and receiver dipoles
($G^\mathrm{ee}_{xx}$), other configurations yield similar results (the choice
is based on the insight that this is the most interesting of all electric
source to electric receiver fields, as can be seen in \cite{GEO.15.Hunziker}
Figures~1 and~2).

@@

\cite{GEO.15.Hunziker} derive the electromagnetic fields in astoundingly simple
equations for all 36 possible source-receiver combinations (electric and
magnetic sources and receivers in three directions $x,y,z$) by finding the
solution for the vertical electric field and then applying the duality
principle and reciprocity to derive all components. The corresponding code
EMmod is published on the SEG website, and in \cite{GEO.16.Hunziker} they
published with iEMmod an inversion routine for it.

@@

Many 1D CSEM codes use the diffusive approximation that is valid for low
frequencies. The appealing part of the derivation of \cite{GEO.15.Hunziker} is
that it models the complete wavefield, hence it is also valid for high
frequencies and therefore wave-phenomena. As an extreme case,
\cite{GEO.15.Hunziker} show a 1D example for ground-penetrating radar with a
center frequency of 250\,MHz. To do so, they calculate 2048 frequencies in the
range of 1\,MHz to 2048\,MHz for the Fast Fourier transform from frequency to
time domain. Here I only calculated the FFT with 850 frequencies from 1\,MHz
to 850\,MHz and then zero-padded up to 2048\,MHz, for EMmod and empymod; tests
have shown that this is sufficient. Figure~\ref{fig:gpr} shows the model with
the survey setup and the four results for EMmod, empymod with QUAD for
relative and absolute tolerance of $10^{-12}$ and $10^{-20}$, respectively,
empymod with a 21\,pt QWE and relative and absolute tolerance of $10^{-10}$ and
$10^{-18}$, respectively, and empymod with FHT with the filter Key09-401,
together with the analytical solution for the arrival of the direct wave (red),
the wave refracted at the surface (yellow), and the wave reflected at the
subsurface interface (green).

@@

Table~\ref{tbl:runtimes} lists the run times for exactly the same models as
\cite{GEO.12.Key} compared in his Table~1: 1\,km water layer, followed by a
1\,km overburden of 1\,$\Omega\,$m, 100\,m reservoir of 100\,$\Omega\,$m, and
lastly 1 or 96 layers of underburden with a resistivity of 1\,$\Omega\,$m;
frequency is 1\,Hz, source depth 990\,m, and receivers are on the sea-bottom;
there are 1, 5, 21, 81, or 321 offsets between 0.5 and 20\,km. In this
comparison, a 9\,pt QWE is compared to a 201\,pt (Key12-201) and a 801\,pt
filter (Anderson82-801), where the absolute and relative tolerance are set so
that the QWE achieves a similar error-level as the filters ($\varepsilon_r =
10^{-6}$ for the standard QWE and $10^{-2}$ for the splined version,
$\varepsilon_a = 10^{-24}$ in both cases). In order to make a fair comparison I
re-run the script from \cite{GEO.12.Key}, and the run times on the test-machine
are roughly twice as fast as in the original paper. Next to it are the results
from empymod. It can be seen from the results that empymod is significantly
faster.%
%
\tabl[btp]{runtimes}%
{Run times comparing the codes from \cite{GEO.12.Key} (Key) and empymod (Wer)
for the same cases as in \cite{GEO.12.Key}, Table~1 (in milliseconds).}{
  \centering
\begin{tabularx}{1.05\textwidth}{ccYYWWYYWWYYWW}
  \toprule
  %
  & & \multicolumn{4}{c}{QWE: 9\,pt}&\multicolumn{4}{c}{FHT: 201\,pt filter}&
  \multicolumn{4}{c}{FHT: 801\,pt filter} \\
  %
  \cmidrule(rl){3-6} \cmidrule(rl){7-10} \cmidrule(rl){11-14}
  %
  & &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Splined} &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Lagged} &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Lagged} \\
  %
  Lay& Off& Wer& Key& Wer& Key& Wer&  Key& Wer& Key& Wer&  Key& Wer& Key\\
  \midrule
  %
    5&   1&   7&  16&   2&  28&   1&   19&   1&  20&   2&   76&   2&  76\\
    5&   5&  13&  69&   4&  29&   3&   92&   1&  24&   7&  365&   3&  82\\
    5&  21&  19& 300&   6&  36&   8&  384&   2&  25&  25& 1522&   3&  83\\
    5&  81&  37&1154&  16&  62&  30& 1475&   2&  28& 100& 5865&   3&  86\\
    5& 321& 108&4581&  58& 165& 124& 5867&   2&  41& 437&23165&   3&  99\\
  \midrule
  100&  21& 118& 463&  13&  51&  92&  604&   8&  37& 304& 2402&  19& 127\\
  100&  81& 296&1792&  23&  75& 337& 2313&   8&  40&1234& 9215&  19& 129\\
  100& 321&1021&7226&  65& 178&1408& 9350&  10&  53&5448&36520&  19& 142\\
  %
  \bottomrule
\end{tabularx}}%
%
Important to note is, however, not the absolute speed of empymod, but the
difference between the various Hankel transform methods. The standard QWE
method, for a similar error level as FHT, is faster than the standard FHT mode
only for many offsets. More importantly, the lagged FHT is generally much
faster than any QWE. QWE is very useful to check the error level of a result.
If many kernel evaluations have to be carried out, and speed is of importance,
the lagged FHT is the preferred method. As \cite{GEO.12.Key} loops over each
wavenumber, the difference between the vectorized and non-vectorized version
can be best seen in the long 801\,pt filter.

@@

Table~\ref{tbl:runtimes2} lists the run times for the same model as before, but
for empymod and Dipole1D using the FHT method with the filter Key09-201, the
default filter in both Dipole1D and empymod. For the regular versions empymod
and Dipole1D perform very similar; empymod is a little faster for the smaller
tests, Dipole1D is slightly faster for the bigger tests. It can be seen from
the results that the parallel optimisation in empymod can result in a
significant speed-up. If the lagged convolution optimisation is chosen,
empymod is significantly faster than Dipole1D. In empymod, additional offsets
come at basically no cost in the lagged version, whereas Dipole1D still uses
more time to calculate more offsets. This is most likely due to the writing of
the output file, that becomes bigger with bigger offsets.
%
\tabl[btp]{runtimes2}
{Run times comparing Dipole1D \citep{GEO.09.Key} and empymod, using the filter
Key09-201 (times in milliseconds). Optimisation: [-] None, [par] parallel,
[spl] lagged FHT.}{ \centering
  \begin{tabularx}{.5\textwidth}{ccAAABB}
  \toprule
  %
     &    & \multicolumn{3}{c}{empymod} & \multicolumn{2}{c}{Dipole1D}\\
  %
  \cmidrule(rl){3-5} \cmidrule(rl){6-7}
  %
  Lay& Off&  - & par & spl &   - & spl\\
  \midrule
  %
    5&   1&    1&   2& 1&    4&  4\\
    5&   5&    3&   3& 2&    6&  5\\
    5&  21&    9&   5& 2&   12&  8\\
    5&  81&   30&  10& 2&   36& 18\\
    5& 321&  124&  34& 2&  130& 56\\
  \midrule
  100&  21&   91&  53& 9&   90& 13\\
  100&  81&  340& 106& 9&  333& 23\\
  100& 321& 1423& 313& 9& 1321& 62\\
  %
  \bottomrule
\end{tabularx}}%
%
Dipole1D was compiled using the free and open-source \emph{gfortran} compiler.
Compiling it with the (proprietary) \emph{ifort} compiler might result in
slightly faster run times. Note that Dipole1D always calculates all six
receiver components, whereas empymod only calculates the required component(s).
On the other hand empymod computes the whole, anisotropic wavefield, whereas
Dipole1D is isotropic and uses the diffusive approximation.

@@

As an exemplary minimal example we calculate a halfspace-model with
air/halfspace interface at $0\,$m; electric, $x$-directed source at
$x_s=y_s=0\,$m and $z_s=1\,$m; electric receiver at $x_r=1000\,$m, $y_r=0\,$m,
$z_r=1\,$m, with azimuth $\theta_r=45^\circ$ and dip $\varphi_r=10^\circ$; and
frequency of $2\,$Hz.

>>>>

%% ~ ADDITIONAL PACKAGES TO GEOPHYSICS.CLS
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\usepackage{color}
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage[USenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
\usepackage[pdftex, final]{hyperref}
\usepackage{upquote}

@@

\definecolor{MyGray}{gray}{0.85}
\newcolumntype{Y}{>{\columncolor{MyGray}\raggedleft\arraybackslash}r}
\newcolumntype{W}{>{\raggedleft\arraybackslash}r}
\newcolumntype{A}{>{\columncolor{MyGray}\raggedleft\arraybackslash}X}
\newcolumntype{B}{>{\raggedleft\arraybackslash}X}

@@

\ms{GEO-2016-0626.R3}

@@

% 1.4 1D codes and licenses
In terms of freely available and open-source code there are a few examples.
\cite{GEO.09.Key} published his forward modeling and inversion code Dipole1D,
written in Fortran. The dipole can be placed anywhere in the stack of layers
and can have arbitrary orientation and dip. Dipole1D computes the electric and
magnetic fields to an electric source, using the FHT method. \cite{GEO.12.Key}
published additional code with his introduction of the
qua\-dra\-ture-with-ex\-tra\-po\-la\-tion method (QWE) and its comparison to
the FHT method, for which he translated some of the Dipole1D-Fortran code to
Matlab. \cite{GEO.15.Hunziker} published their code EMmod (Fortran and C), in
which the source and receivers can also be placed anywhere in the stack of
layers. They use a 61\,pt Gauss-Kronrod quadrature for the Hankel transform.

@@

To my knowledge, empymod is the first code that is freely available which
calculates the full wavefield for a layered-earth model with vertical isotropy
and has different types of Hankel transforms (QWE, QUAD, FHT) and various
methods for the Fourier transform (FFTLog, sine/cosine-filters, QWE, FFT). An
interesting use-case for empymod is therefore comparison studies of the
different methods, up to building hybrid inversion schemes. The code is
published under the lax permissive \emph{Apache Version 2.0 license}, which
makes it available to everyone for free, even for commercial purposes. It might
hence prove to be valuable for students and professionals alike, and it is
valuable for the objectives of reproducible research. It is written in Python,
a modern, cross-platform, free and open-source programming language. With its
scientific libraries, mainly the numeric and scientific modules NumPy and
SciPy, it creates an extremely powerful numerical calculation stack: Even
though it is an interpreted language it can be very fast, as NumPy/SciPy use
under the hood routines in Fortran or in C (using for instance the BLAS/Lapack
libraries) for the linear algebra computations, and Python is merely the glue.
The comparisons show that empymod is as fast as the existing codes. The code is
vectorized (avoiding loops) as much as possible, which improves computation.
The most time-consuming calculations have furthermore a flag to run
parallelized. However, on a larger scale such as an inversion routine one would
probably want to parallelize the kernel calls instead of the kernel itself. The
code is hosted on GitHub, which makes it easy for anyone to improve and
contribute to the code. It comes with an extensive testing suite which is run
automatically on each commit, guaranteeing robustness, and helps to avoid
regression bugs. Installation guidelines, documentation, and examples of its
usage can be found on its website at \url{https://empymod.github.io}.
\cite{TLE.17.Werthmuller} gives a tutorial-style introduction to CSEM modeling
using empymod.

@@

The wavenumber-domain calculation in \emph{kernel} follows
\cite{GEO.15.Hunziker}, and calculates as such the complete wavefield for a
layered VTI model. The code makes no assumptions about the model: Source and
receiver can be placed anywhere in the model, including first and last layer,
and you have to define if the first layer is air or not. Bipoles can cross
layer boundaries. Depths, frequencies, and source-receiver configuration have
to be defined, and each layer is characterized with its horizontal resistivity
$\rho_\textrm{h}$, its electrical anisotropy $\lambda$, where $\lambda =
\sqrt{\rho_\textrm{v}/\rho_\textrm{h}}$, its horizontal and vertical magnetic
permeabilities $\mu_\mathrm{h}$ and $\mu_\mathrm{v}$, and the horizontal and
vertical electric permittivities $\epsilon_\mathrm{h}$ and
$\epsilon_\mathrm{v}$. The main difference between EMmod and empymod is the
Hankel transform, and a few additional fundamental differences: EMmod uses 2nd
order Bessel functions of the first kind $J_2$. Published FHT filters generally
provide coefficients for 0th and 1st order Bessel functions $J_0$, $J_1$ only.
Consequently, the recurrence relation
%
\begin{equation}
  J_2(kr) = \frac{2}{kr}J_1(kr) - J_0(kr)
  \label{eq:j2}
\end{equation}
%
is used, where $k$ and $r$ are the wavenumber and space-domain parameters,
respectively. Other differences are that empymod includes Fourier transforms,
hence time-domain calculation, and can model arbitrary rotated, finite bipoles.

@@

The most import part of \cite{GEO.12.Key} is the introduction of a new
quadrature algorithm to geophysics, named
qua\-dra\-ture-with-ex\-tra\-po\-la\-tion (QWE). QWE is a fast quadrature
method using the Shanks transformation \citep{JMP.55.Shanks} computed with
Wynn's epsilon algorithm \citep{MC.56.Wynn}. The advantage of quadrature over
filters is the ability to estimate the error. QWE continues until the absolute
error, estimated by the difference of subsequent iterations $n$, satisfies the
inequality
%
\begin{equation}
  |S^*_n-S^*_{n-1}| \le \varepsilon_r|S^*_n| + \varepsilon_a\ ,
  \label{eq:err}
\end{equation}
%
where $S^*$ is the extrapolated result, and $\varepsilon_r$, $\varepsilon_a$
are the relative and absolute tolerance, respectively.

Whereas for the Hankel transform three methods are implemented with QWE, FHT,
and QUAD, a standard adaptive quadrature using QAGSE from the Fortran QUADPACK
library, for the Fourier transform there are four methods implemented: QWE, FHT
with the sine and cosine filters, the standard FFT, and FFTLog
\citep{RAS.00.Hamilton}. The logarithmic fast Fourier transform FFTLog is ideal
for this operation, as generally a wide range of frequencies is required to go
from the frequency to the time domain. The FFTLog can yield faster results than
the QWE method and more precise than the FHT results, specifically for land
impulse responses at early times.

Calculation time can be reduced by using the splined version of the QWE and the
splined and lagged versions of the FHT. In addition, all time-consuming
calculations are set up to be able to run in parallel, with the help of the
Python-module numexpr. This can significantly speed up calculations if you run
big models with many layers and for many offsets and frequencies.  However, if
you include empymod in an inversion scheme then it might be better to
parallelize the calls to empymod, instead of empymod itself.

@@

The run-time comparison tests were run on a Lenovo ThinkCentre running Ubuntu
16.04 64-bit, with 8\,GB of memory and an Intel Core i7-4770 CPU @ 3.40GHz x 8.
Comparing run times is always a difficult task, and between different languages
even more, here between Python (empymod), Fortran (EMmod, Dipole1D), and Matlab
(scripts from \cite{GEO.12.Key}). However, they do serve as comparison. I used
the Matlab-\texttt{timing} and Python-\texttt{timeit} functions, which behave
very similar. It is probably expected today, but still worth mentioning, that
the calculation is carried out in double precision.

@@

\cite{PIER.10.Slob} published analytical frequency- and time-domain solutions
for the diffusive electric field in a VTI half-space, where source and receiver
can be placed anywhere in the half-space. As an example, I use the same model
as \cite{GEO.15.Hunziker} in their Figures~1 and~2: A half-space with
horizontal resistivity $\rho_\mathrm{h} = 1/3\,\Omega$\,m, anisotropy $\lambda
= \sqrt{10}$, and for frequency $f = 0.5\,$Hz. The source is located
horizontally at the origin ($x_\mathrm{s} = y_\mathrm{s} = 0\,$m) at a depth of
150\,m, and the depth of the receivers is 200\,m. The field is calculated on a
regular grid with a spacing of 10\,m. Figure~\ref{fig:analytical} shows the
analytical amplitude and phase results for the first quadrant (the other
quadrants are simply symmetric copies of it).
%
\plot*[btp]{analytical}{width=\textwidth}{Analytical solution for a half-space
  with $\rho_\mathrm{h} = 1/3\,\Omega$\,m, $\lambda = \sqrt{10}$, $f =
  0.5\,$Hz, $z_\mathrm{s} = 150\,$m, and $z_\mathrm{r} = 200\,$m, calculated on
  a regular grid with a spacing of 10 meters; (a) amplitude and (b) phase.}
%
The figure shows the result for x-directed electric source and receiver dipoles
($G^\mathrm{ee}_{xx}$), other configurations yield similar results (the choice
is based on the insight that this is the most interesting of all electric
source to electric receiver fields, as can be seen in \cite{GEO.15.Hunziker},
Figures~1 and~2).

@@

\cite{GEO.15.Hunziker} derive the electromagnetic fields in astoundingly simple
equations for all 36 possible source-receiver combinations (electric and
magnetic sources and receivers in three directions $x$, $y$, $z$) by finding
the solution for the vertical electric field and then applying the duality
principle and reciprocity to derive all components. The corresponding code
EMmod is published on the SEG website, and in \cite{GEO.16.Hunziker} they
published with iEMmod an inversion routine for it.

@@

Many 1D CSEM codes use the diffusive approximation that is valid for low
frequencies. The appealing part of the derivation of \cite{GEO.15.Hunziker} is
that it models the complete wavefield, hence it is also valid for high
frequencies and therefore wave-phenomena. As an extreme case,
\cite{GEO.15.Hunziker} show a 1D example for ground-penetrating radar with a
center frequency of 250\,MHz. To do so, they calculate 2048 frequencies in the
range of 1\,MHz to 2048\,MHz for the Fast Fourier transform from frequency to
time domain. Here I only calculated the FFT with 850 frequencies from 1\,MHz to
850\,MHz and then zero-padded up to 2048\,MHz, for EMmod and empymod; tests
have shown that this is sufficient. Figure~\ref{fig:gpr} shows the model with
the survey setup and the four results for EMmod, empymod with QUAD for relative
and absolute tolerance of $10^{-12}$ and $10^{-20}$, respectively, empymod with
a 51\,pt QWE and relative and absolute tolerance of $10^{-8}$ and $10^{-15}$,
respectively, and empymod with FHT with the filter Key09-401, together with the
analytical solution for the arrival of the direct wave (red), the wave
refracted at the surface (yellow), and the wave reflected at the subsurface
interface (green).

@@

Table~\ref{tbl:runtimes} lists the run times for exactly the same models as
\cite{GEO.12.Key} compared in his Table~1: 1\,km water layer, followed by a
1\,km overburden of 1\,$\Omega\,$m, 100\,m reservoir of 100\,$\Omega\,$m, and
lastly 1 or 96 layers of underburden with a resistivity of 1\,$\Omega\,$m;
frequency is 1\,Hz, source depth 990\,m, and receivers are on the sea-bottom;
there are 1, 5, 21, 81, or 321 offsets between 0.5 and 20\,km. In this
comparison, a 9\,pt QWE is compared to a 201\,pt (Key12-201) and a 801\,pt
filter (Anderson82-801), where the absolute and relative tolerance are set so
that the QWE achieves a similar error-level as the filters ($\varepsilon_r =
10^{-6}$ for the standard QWE and $10^{-2}$ for the splined version,
$\varepsilon_a = 10^{-24}$ in both cases). In order to make a fair comparison I
re-run the script from \cite{GEO.12.Key}, and the run times on the test-machine
are roughly twice as fast as in the original paper. Next to it are the results
from empymod. It can be seen from the results that empymod is significantly
faster.%
%
\tabl[btp]{runtimes}{Run times comparing the codes from \cite{GEO.12.Key} (Key)
  and empymod (Wer) for the same cases as in \cite{GEO.12.Key}, Table~1 (in
  milliseconds).}{
  \centering
\begin{tabularx}{1.05\textwidth}{ccYYWWYYWWYYWW}
  \toprule
  %
  & & \multicolumn{4}{c}{QWE: 9\,pt}&\multicolumn{4}{c}{FHT: 201\,pt filter}&
  \multicolumn{4}{c}{FHT: 801\,pt filter} \\
  %
  \cmidrule(rl){3-6} \cmidrule(rl){7-10} \cmidrule(rl){11-14}
  %
  & &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Splined} &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Lagged} &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Lagged} \\
  %
  Lay& Off& Wer& Key& Wer& Key& Wer&  Key& Wer& Key& Wer&  Key& Wer& Key\\
  \midrule
  %
    5&   1&   7&  16&   2&  28&   1&   19&   1&  20&   2&   76&   2&  76\\
    5&   5&  13&  69&   4&  29&   3&   92&   1&  24&   7&  365&   3&  82\\
    5&  21&  19& 300&   6&  36&   8&  384&   2&  25&  25& 1522&   3&  83\\
    5&  81&  37&1154&  16&  62&  30& 1475&   2&  28& 100& 5865&   3&  86\\
    5& 321& 108&4581&  58& 165& 124& 5867&   2&  41& 437&23165&   3&  99\\
  \midrule
  100&  21& 118& 463&  13&  51&  92&  604&   8&  37& 304& 2402&  19& 127\\
  100&  81& 296&1792&  23&  75& 337& 2313&   8&  40&1234& 9215&  19& 129\\
  100& 321&1021&7226&  65& 178&1408& 9350&  10&  53&5448&36520&  19& 142\\
  %
  \bottomrule
\end{tabularx}}%
%
Important to note is, however, not the absolute speed of empymod, but the
difference between the various Hankel transform methods. The standard QWE
method, for a similar error level as FHT, is faster than the standard FHT mode
only for many offsets. More importantly, the lagged FHT is generally much
faster than any QWE. QWE is very useful to check the error level of a result.
If many kernel evaluations have to be carried out, and speed is of importance,
the lagged FHT is the preferred method. As \cite{GEO.12.Key} loops over each
wavenumber, the difference between the vectorized and non-vectorized version
can be best seen in the long 801\,pt filter.

@@

Table~\ref{tbl:runtimes2} lists the run times for the same model as before, but
for empymod and Dipole1D using the FHT method with the filter Key09-201, the
default filter in both Dipole1D and empymod. For the regular versions empymod
and Dipole1D perform very similar; empymod is a little faster for the smaller
tests, Dipole1D is slightly faster for the bigger tests. It can be seen from
the results that the parallel optimisation in empymod can result in a
significant speed-up. If the lagged convolution optimisation is chosen, empymod
is significantly faster than Dipole1D. In empymod, additional offsets come at
basically no cost in the lagged version, whereas Dipole1D still uses more time
to calculate more offsets. This is most likely due to the writing of the output
file, that becomes bigger with bigger offsets.
%
\tabl[btp]{runtimes2}{Run times comparing Dipole1D \citep{GEO.09.Key} and
  empymod, using the filter Key09-201 (times in milliseconds). Optimisation:
  [-] None, [par] parallel, [spl] lagged FHT.}{
  \centering
  \begin{tabularx}{.5\textwidth}{ccAAABB}
  \toprule
  %
     &    & \multicolumn{3}{c}{empymod} & \multicolumn{2}{c}{Dipole1D}\\
  %
  \cmidrule(rl){3-5} \cmidrule(rl){6-7}
  %
  Lay& Off&  - & par & spl &   - & spl\\
  \midrule
  %
    5&   1&    1&   2& 1&    4&  4\\
    5&   5&    3&   3& 2&    6&  5\\
    5&  21&    9&   5& 2&   12&  8\\
    5&  81&   30&  10& 2&   36& 18\\
    5& 321&  124&  34& 2&  130& 56\\
  \midrule
  100&  21&   91&  53& 9&   90& 13\\
  100&  81&  340& 106& 9&  333& 23\\
  100& 321& 1423& 313& 9& 1321& 62\\
  %
  \bottomrule
\end{tabularx}}%
%
Dipole1D was compiled using the free and open-source \emph{gfortran} compiler.
Compiling it with the (proprietary) \emph{ifort} compiler might result in
slightly faster run times. Note that Dipole1D always calculates all six
receiver components, whereas empymod only calculates the required component(s).
On the other hand empymod computes the whole, anisotropic wavefield, whereas
Dipole1D is isotropic and uses the diffusive approximation.

@@

As a minimal example we calculate a half-space model with air/half-space
interface at $0\,$m; electric, $x$-directed source at $x_s=y_s=0\,$m and
$z_s=1\,$m; electric receiver at $x_r=1000\,$m, $y_r=0\,$m, $z_r=1\,$m, with
azimuth $\theta_r=45^\circ$ and dip $\varphi_r=10^\circ$; and frequency of
$2\,$Hz.

\begin{verbatim}
>>> from empymod import bipole
>>> em = bipole(
>>>     src = [0, 0, 1, 0, 0],
>>>     rec = [1000, 0, 1, 45, 10],
>>>     depth = 0,         # interface
>>>     res = [2e14, 100], # air, half-space
>>>     freqtime = 2,      # frequency
>>>     verb = 1)          # verbosity
>>> print('%0.4e, %0.4ei' % (em.real, em.imag))
2.2042e-08, -7.1539e-10i
\end{verbatim}
